Analyzing Fairness: Race, Recidivism, and the COMPAS Algorithm

### 2022-04-14 Paper_Draft2/Draft2.tex

The data we are using for this project is the COMPAS General Recidivism
Risk Scores data set from the AI Fairness 360 (AIF360) toolkit
\citep{aif360-oct-2018}, which does the same initial pre-processing as
ProPublica. The raw data has 6,167 rows and each row represents an
arrest charge for a defendant. AIF360's COMPAS data includes the
defendant's age, race, sex, what they were charged with, and whether or
not the defendant ultimately recidivated within a two-year period after
their arrest. For the purposes of our project, which endeavors to
evaluate the differing effects of the COMPAS algorithm on white
defendants versus Black defendants, we have filtered the data to only
include individuals whose race is listed as Caucasian or
African-American. Our data therefore has 5,723 rows (Figure
\ref{fig:table snip}), with the below distributions of race (Figure
\ref{fig:race plot}), age (Figure \ref{fig:age plot}), and two year
recidivism rate (Figures \ref{fig:recid plot} \&
\ref{fig:recid race plot}).

@@

{\centering \includegraphics[width=500px,height=300px]{../images/table_snippet_new}

@@

{\centering \includegraphics[width=500px,height=300px]{../images/race_bar_plot}

@@

{\centering \includegraphics[width=500px,height=300px]{../images/race_age_plot}

@@

{\centering \includegraphics[width=500px,height=300px]{../images/recid_bar_plot}

@@

{\centering \includegraphics[width=500px,height=300px]{../images/race_recid_plot}

@@

>>>>

The data we are using for this project is the COMPAS General Recidivism
Risk Scores data set from the AI Fairness 360 (AIF360) toolkit
\citep{aif360-oct-2018}, which does the same initial pre-processing as
ProPublica. The raw data has 6,167 rows and each row represents an
arrest charge for a defendant. AIF360's COMPAS data includes the
defendant's age, race, sex, what they were charged with, and whether or
not the defendant ultimately recidivated within a two-year period after
their arrest. For the purposes of our project, which endeavors to
evaluate the differing effects of the COMPAS algorithm on white
defendants versus Black defendants, we have filtered the data to only
include individuals whose race is listed as Caucasian or
African-American. Our data therefore has 5,723 rows (Figure
\ref{fig:table snip}), with the below distributions of race (Figure
\ref{fig:race plot}), age (Figure \ref{fig:age plot}), and two year
recidivism rate (Figures \ref{fig:recid plot} \&
\ref{fig:recid race plot} and Table \ref{tab:recid table}).

@@

{\centering \includegraphics[width=1\linewidth]{../images/table_snippet}

@@

{\centering \includegraphics[width=1\linewidth]{../images/race_bar_plot_new}

@@

{\centering \includegraphics[width=1\linewidth]{../images/age_race_plot_new}

@@

{\centering \includegraphics[width=1\linewidth]{../images/recid_bar_plot_new}

@@

{\centering \includegraphics[width=1\linewidth]{../images/race_recid_bar_plot_new}

@@

\begin{longtable}[]{@{}lll@{}}
\caption{This table shows the distribution of recidivism by race
\label{tab:recid table}}\tabularnewline
\toprule
Two Year Recidivism by Race & Recidivated & Survived\tabularnewline
\midrule
\endfirsthead
\toprule
Two Year Recidivism by Race & Recidivated & Survived\tabularnewline
\midrule
\endhead
African American & 1661 & 1512\tabularnewline
Caucasian & 822 & 1278\tabularnewline
\bottomrule
\end{longtable}

### 2022-04-14 final_metrics_analysis.Rmd

```{r py setup, include=FALSE}
library(reticulate)
use_condaenv("anaconda3")
```

@@

purple = '#9d0677'
green = '#30875c'
workshop_palette = [purple, green]
df_viz.head()
```

@@

```{python exploratory race bar plot}
# barplot of race
sns.countplot(y="race", data=df_viz, palette=["#30875c"])
plt.title('Race Distribution (White and Black individuals only)')
```

```{python exploratory age by race plot}
# age distribution by race
#ax = sns.kdeplot(x="age", hue="race", data=df_viz);
#kdeline = ax.lines[0]
#mean_fem = df_viz.groupby('race').age.median()[0]
#xs = kdeline.get_xdata()
#ys = kdeline.get_ydata()
#height_fem = np.interp(mean_fem, xs, ys)
#ax.vlines(mean_fem, 0, height_fem, color=purple, ls=':')

#mean_m = df_viz.groupby('race').age.median()[1]
#height_m = np.interp(mean_m, xs, ys)
#ax.vlines(mean_m, 0, height_m, color=green, ls=':')

#print(df_viz.groupby('race').age.median())

#plt.title('Distribution of Age by Race')
```

>>>>

```{r py setup, include=FALSE}
library(reticulate)
```

@@

purple = '#9d0677'
green = '#30875c'
orange = '#E7881E'
blue = '#20A4CF'
workshop_palette = [purple, green]
df_viz.head()
```

@@

```{python exploratory race bar plot}
# barplot of race
sns.countplot(x="race", data=df_viz, palette=[blue, orange])
plt.title('Race Distribution (White and Black individuals only)')
```

```{python exploratory age by race plot}
# age distribution by race
ax = sns.kdeplot(x="age", hue="race", data=df_viz);
kdeline = ax.lines[0]
mean_fem = df_viz.groupby('race').age.median()[0]
xs = kdeline.get_xdata()
ys = kdeline.get_ydata()
height_fem = np.interp(mean_fem, xs, ys)
ax.vlines(mean_fem, 0, height_fem, color=blue, ls=':')

mean_m = df_viz.groupby('race').age.median()[1]
height_m = np.interp(mean_m, xs, ys)
ax.vlines(mean_m, 0, height_m, color=orange, ls=':')

print(df_viz.groupby('race').age.median())

plt.title('Distribution of Age by Race')
```
